\textbf{Too Beautiful to be Fake: Attractive Faces are Less Likely to be Judged as Artificially Generated}

### 2022-10-22 submission/manuscript.Rmd
discussion

This study aimed at investigating the effect of facial ratings (attractiveness, beauty, trustworthiness and familiarity) on simulation monitoring, i.e., on the belief that a stimulus was artificially generated. Most strikingly, despite all the stimuli being real faces from the same database, all participants believed (to high degrees of confidence) that a significant proportion of them were fake. This finding hence attests to the effectiveness of our instructions, highlighting the current expectations regarding CGI technology, as well as the volatility of our sense of reality. It underlines the strong impact of prior expectations and information on reality beliefs. In fact, stimuli-related and participant-related characteristics accounted together for less than 20% of the beliefs variance, suggesting that a large part of it is associated with other subjective processes.

@@

Contrary to our hypothesis, we did not find familiarity to be significantly related to simulation monitoring decisions. Interestingly, there were significant linear relationships between familiarity and confidence judgements for males only, where familiarity increased the confidence of reality beliefs. While this alludes to confidence judgements in face simulation monitoring decisions being more stable in females [@mata2013metacognitive; @klein2018metacognitive], the present literature regarding gender differences in face recognition confidence has been mixed. Namely, previous reports of females demonstrating superior facial memory [@mishra2019gender; @lewin2002sex; @sommer2013sex] and higher metacognitve awareness [@kaplan2012effect], have been contested by recent findings of males' overconfidence in face recall [@bailey2021gender; @herbst2020gender]. Furthermore, it should be noted that the present study's distribution of familiarity ratings was strongly skewed, and only a low number of pictures was rated as highly familiar. As such, future studies should clarify this point by experimentally manipulating familiarity, for instance by modulating the amount of exposure to items before querying the simulation monitoring judgements.

@@

Regarding the role of inter-individual characteristics in simulation monitoring tendencies, we found higher scores of honesty-humility, a trait related to an increased risk perception and aversion [@levidi2022understanding; @weller2012honest], to be related to a lower confidence in simulation monitoring judgements. Notably, greater narcissistic tendencies in dimensions such as acclaim seeking and manipulativeness was associated with a higher number of faces judged as real. This is in line with recent research which found narcissists were less likely to engage in analytical reasoning strategies such as reflective thinking [@littrell2020overconfidently; @ahadzadeh2021social], and are more cognizant of external stimuli [@carolan2017searching; @grapsas2020and; @eddy2021self].

Moreover, putting the significant positive links between narcissistic acclaim seeking and confidence judgements in perspective with the negative correlation between honesty-humility and narcissism [@hodson2018dark], we confirm previous evidence of the relationship between narcissistic grandiosity and over-confidence in decision-making [@campbell2004narcissism; @chatterjee2017master; @brunell2017narcissism; @o2021grandiose]. Although an inverse effect was found for the narcissistic facet of authoritativeness, we interpret this correlation as related to a higher response assertiveness. Taken together, these results suggest that participants with low humility and high recognition desires are more confident in their judgement regarding the real or fake nature of ambiguous stimuli. Alternatively, participants with opposite traits might perceive a higher risk in the decision-making process and its potential consequences (e.g., being seen as bad at the task at hand), resulting in more conservative confidence ratings.

@@

Our findings suggest - though with weak significance - a positive link between paranoid ideation and the belief that the stimuli were real. Given previous reports that people with higher levels of paranoia are more sensitive to cues of social threat [@fornells2015people; @freeman2003can; @king2017paranoia], it could be that individuals with paranoid traits confer greater saliency and emotionality to observed faces, hence increasing perceptions of its realness. This hypothesis, if confirmed by future studies, would be in line with previous findings that persecutory delusions are predicted by a greater sense of presence in a VR environment populated with virtual characters [@freeman2005psychology].

@@

<!-- given the known positive correlations between paranoia, social anxiety and uncertainty intolerance [@morrison2014moderating; @counsell2017intolerance; @barnby2020paranoia] -->

@@

<!-- However, it is important to note that all of the aforementioned dispositional traits did only influence the confidence judgements, but not the simulation monitoring bias *per se*. -->

@@

On a methodological level, although the order of presentation of the facial images was randomized to reduce effects of adaptation, participants were more confident in their judgements for faces perceived as real following a shorter re-exposure delay. This may be because shorter re-exposure delays, as opposed to longer delays, led to the faces being better remembered and appearing more familiar, thereby triggering self-referential and autobiographical memory processing during the repeated display [@abraham2009reality; @gobbini2013prioritized; @taylor2009neural]. Indeed, this finding is consistent with studies in which fictional stimuli that were associated with familiarity up-regulated emotions, biasing its salience and perceived realness [@sperduti2016paradox; @makowski2017being]. However, if that was the case, we would expect shorter re-exposure delays to impact the decision bias as well towards reality, rather than simply the confidence. Future studies should further investigate the modulatory effects of types and degrees of familiarity on perceived realness judgements.

@@

Understanding the factors driving the reality beliefs pertaining to ambiguous stimuli is becoming increasingly pertinent given the rapid technological advances. The current experimental paradigm required participants to judge the realness of faces they had prior exposure to (which was done to prevent reality judgements from influencing the other ratings). Although the effect of re-exposure delay was negligible, the potential bias induced by face familiarity (as compared to judging completely new items) cannot be discarded. Future studies could examine that by incorporating novel face images or increasing the duration of the re-exposure delay.

The aim of the present study was to examine whether a subset of specific characteristics, in particular face attractiveness, significantly influences our simulation monitoring decisions. Notably, we found faces rated as attractive to be perceived as more real, with a possible sexual dimorphism affecting the shape of the relationship. We also found that inter-individual traits, such as narcissistic acclaim-seeking and manipulativeness, as well as persecutory ideation, were related to a systematic bias towards beliefs that the stimuli were real or fake. However, it has to be noted that the magnitude of the effects found was relatively small, suggesting that the facial features measured in the study were not the key determinants of simulation monitoring. Hence, beyond exploring new potential mechanisms, future studies should include a more thorough debriefing to try to capture what conscious strategies (if any) the participants used (e.g., focusing on some features of the stimulus - like hair or eyes in the case of faces) to guide their reality beliefs.

>>>>

This study aimed at investigating the effect of facial ratings (attractiveness, beauty, trustworthiness and familiarity) on simulation monitoring, i.e., on the belief that a stimulus was artificially generated. Most strikingly, despite all the stimuli being real faces from the same database, all participants believed (to high degrees of confidence) that a significant proportion of them were fake. This finding not only attests to the effectiveness of our instructions, but highlights the current levels of expectation regarding CGI technology. The strong impact of prior expectations and information on reality beliefs demonstrated here underlines the volatility of our sense of reality. In fact, stimuli-related and participant-related characteristics accounted together for less than 20% of the beliefs variance, suggesting a large contribution of other subjective processes.

@@

Contrary to our hypothesis, we did not find familiarity to be significantly related to simulation monitoring decisions. Interestingly, there were significant linear relationships between familiarity and confidence judgements for males only, where familiarity increased the confidence of reality beliefs. Although the familiarity measure was not a "recognition" measure, evidence from studies pertaining to the latter could be linked, reporting better face memory for females [@mishra2019gender; @lewin2002sex; @sommer2013sex], as well as an overconfidence in face recall for males [@bailey2021gender; @herbst2020gender]. However, it should be noted that the present study's distribution of familiarity ratings was strongly skewed, and only a low number of pictures was rated as highly familiar. As such, future studies should clarify this point by experimentally manipulating familiarity, for instance by modulating the amount of exposure to items before querying the simulation monitoring judgements.

@@

Regarding the role of inter-individual characteristics in simulation monitoring tendencies, we found higher scores of honesty-humility - a trait related to an increased risk perception and aversion [@levidi2022understanding; @weller2012honest] - to be related to a lower confidence in simulation monitoring judgements. Notably, greater narcissistic tendencies in dimensions such as acclaim seeking and manipulativeness were associated with a higher number of faces judged as real. This is in line with recent research which found people with narcissism to be less likely to engage in analytical reasoning strategies such as reflective thinking [@littrell2020overconfidently; @ahadzadeh2021social], and to be more vigilant and attentive to external stimuli [@carolan2017searching; @grapsas2020and; @eddy2021self].

Moreover, putting the significant positive links between narcissistic acclaim seeking and confidence judgements in perspective with the negative correlation between honesty-humility and narcissism [@hodson2018dark], we confirm previous evidence regarding the relationship between narcissistic grandiosity and over-confidence in decision-making [@campbell2004narcissism; @chatterjee2017master; @brunell2017narcissism; @o2021grandiose]. Although an inverse effect was found for the narcissistic facet of authoritativeness, we interpret this relationship as related to a higher response assertiveness. Taken together, these results suggest that participants with low humility and high recognition desires are more confident in their judgement regarding the real or fake nature of ambiguous stimuli. Alternatively, participants with opposite traits might perceive a higher risk in the decision-making process and its potential consequences (e.g., being seen as bad at the task at hand), resulting in more conservative confidence ratings.

@@

Our findings suggest - though with weak significance - a positive link between paranoid ideation and the tendency to believe that the stimuli were real. Given previous reports that people with higher levels of paranoia are more sensitive to cues of social threat [@fornells2015people; @freeman2003can; @king2017paranoia], it is plausible that paranoid traits confer greater saliency and emotionality to observed faces, hence increasing perceptions of its realness. This hypothesis, if confirmed by future studies, would be in line with previous findings that persecutory delusions are predicted by a greater sense of presence in VR environments populated with virtual characters [@freeman2005psychology].

@@

@@

@@

On a methodological level, although the order of presentation of the facial images was randomized to reduce effects of adaptation, participants were more confident in their judgements for faces perceived as real following a shorter re-exposure delay. Such shorter durations could be associated with the faces being better remembered and appearing more familiar, thereby triggering self-referential and autobiographical memory processing during the repeated display [@abraham2009reality; @gobbini2013prioritized; @taylor2009neural]. Indeed, this finding is consistent with studies in which fictional stimuli that were associated with familiarity up-regulated emotions, biasing its salience and perceived realness [@sperduti2016paradox; @makowski2017being]. However, if that was the case, we would expect shorter re-exposure delays to impact the decision bias as well towards reality, rather than simply the confidence. Future studies should further investigate the modulatory effects of types and degrees of familiarity on perceived realness judgements.

<!-- Limitations -->

Several limitations have to be noted. The current experimental paradigm required participants to judge the realness of faces they had prior exposure to (which was done to prevent reality judgements from influencing the other ratings). Although the effect of re-exposure delay was negligible, the potential bias induced by face familiarity (as compared to judging completely new items) cannot be discarded. Future studies could examine that by incorporating novel face images or increasing the duration of the re-exposure delay. Moreover, the magnitude of the effects found in the study was relatively small, suggesting that the facial features measured in the study were not the key determinants of simulation monitoring. Hence, beyond exploring new potential mechanisms, future studies should include a more thorough debriefing to try to capture what conscious strategies (if any) the participants used (e.g., focusing on some features of the stimulus - like hair or eyes in the case of faces) to guide their reality beliefs.

@@

In summary, the aim of the present study was to examine whether a subset of specific characteristics, in particular face attractiveness, significantly influences our simulation monitoring decisions. Notably, we found faces rated as attractive to be perceived as more real, with a possible sexual dimorphism affecting the shape of the relationship. We also found that inter-individual traits, such as narcissistic acclaim-seeking and manipulativeness, as well as persecutory ideation, were related to a systematic bias towards beliefs that the stimuli were real or fake. We hope subsequent works to further our understanding of the factors driving the reality beliefs pertaining to ambiguous stimuli, an question becoming increasingly pertinent given the rapid technological advances.
